---
title: "i-Tree Data Exploration"
author: "Matt Fuller"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: yes
    number_sections: yes
    toc_float: yes
    code_folding: hide
  pdf_document:
    toc: yes
  word_document:
    toc: yes
always_allow_html: yes
---

Initial set up and R package loading.  

```{r setup, message=FALSE}
library(sf)
library(tidyverse)
library(DT)
library(leaflet)
library(htmltools)
library(viridis)

birditree_sitejoin_key <- readRDS(file = "../data/iTree_Bird_site_key.RDS")

knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '')
```

# Read in i-Tree Data  

Data were previously cleaned and organized into the below data files that are read in. See the "00_data_cleaning.Rmd" script for details on modifications made to the original data files. In addition to these cleaning steps, below we filter to just the i-Tree plots that are paired with bird sites for an initial data exploration and summary.    

```{r}
itree_plot_df <- readRDS(file = "../data/itree/itree_plot_data_organized.RDS") |>
  left_join(select(birditree_sitejoin_key, itreeplotID, BirdPairedPlotNum = PlotNum), 
            by = "itreeplotID") |>  filter(!is.na(BirdPairedPlotNum)) |>
  select(-BirdPairedPlotNum)
itree_tree_df <- readRDS(file = "../data/itree/itree_tree_data_organized.RDS") |>
  left_join(select(birditree_sitejoin_key, itreeplotID, BirdPairedPlotNum = PlotNum), 
            by = "itreeplotID") |>  filter(!is.na(BirdPairedPlotNum)) |>
  select(-BirdPairedPlotNum)
itree_shrub_df <- readRDS(file = "../data/itree/itree_shrub_data_organized.RDS") |>
  left_join(select(birditree_sitejoin_key, itreeplotID, BirdPairedPlotNum = PlotNum), 
            by = "itreeplotID") |>  filter(!is.na(BirdPairedPlotNum)) |>
  select(-BirdPairedPlotNum)
itree_benefits_df <- 
  readRDS(file = "../data/itree/itree_benefits_data_organized.RDS") |>
  left_join(select(birditree_sitejoin_key, itreeplotID, BirdPairedPlotNum = PlotNum), 
            by = "itreeplotID") |>  filter(!is.na(BirdPairedPlotNum)) |>
  select(-BirdPairedPlotNum)
```



# Map of plots  

To map the plots around Santo Domingo, we can convert the text data with lat-long coordinates into simple features (sf) spatial points data frames. Each sector is has a specified color:  
- Ciudad Nueva: black  
- Gazcue: purple  
- San Carlos: pink  
- Zona Colonial : orange  


```{r}
# create 'sf' R object for plot data to use with 'leaflet' maps
plot_sf <- itree_plot_df |> sf::st_as_sf(coords = c("long", "lat") )
```



```{r warning = FALSE}
# Set up basemap tiles for various background maps
basemap <- leaflet() |>
  # add background maps
  addProviderTiles("Esri.WorldTopoMap", group = "Esri.WorldTopoMap") |>
  addProviderTiles("Esri.WorldImagery", group = "Esri.WorldImagery") |>  
  addProviderTiles("OpenStreetMap.Mapnik", group = "OpenStreetMap.Mapnik") |>
  # add a layers control
  addLayersControl(baseGroups = c("Esri.WorldTopoMap",
                                  "Esri.WorldImagery",
                                  "OpenStreetMap.Mapnik"),
  # position background map selection icon in the top-left part of map
  position = "topleft")
```

Generate searchable map with i-Tree plots.  

```{r warning = FALSE}
# create palette for sector markers
sector_cols <- colorFactor(palette = viridis(4, option = "A", end = 0.85), 
                           levels = factor(plot_sf$SectorID,
                                           levels = c("CN","GA","SC","ZC")) )

# Map the plots for each sector.  
(SD_plot_leaflet <- basemap |>
    addCircleMarkers(data = plot_sf,
                     popup = ~paste0("Sector: ", SectorID,";  Plot: ", PlotNum, "<br>",
                                     "Address: ", Address, "<br>",
                                     "Notes: ", Notes), 
                     radius = 4, opacity = 1,
                     color = ~ sector_cols(SectorID) ) 
)
```


<br> 

## Percent coverage of plots  

Plot data exported from i-Tree were accompanied by summary statistics for percent coverage by trees or shrubs for each plot. Additionally, there was a field for percent coverage plantable in each plot. These coverages were summarized in a figure to compare each sectors' mean percent coverage for trees, shrubs, and plantable area.  


```{r plotsectorcoverages, fig.width=6.5, fig.height=3}
tree_col <- "#1b9e77"
shrub_col <- "#d95f02"
plantable_col <- "#7570b3"
pt_size = 3

(plot_pctcoverage <- 
    ggplot(data = itree_plot_df |> reframe(.by = SectorID, 
                                           percTree = mean(percTree),
                                           percShrub = mean(percShrub),
                                           percPlantable = mean(percPlantable)) ) + 
    # Tree Points
    geom_point(aes(x = SectorID, y = `percTree`, 
                   color = "Tree", shape = "Tree" ), size = pt_size) +
    # Shrub Points
    geom_point(aes(x = SectorID, y = `percShrub`, 
                   color = "Shrub", shape = "Shrub" ), size = pt_size) +
    # Plantable Points
    geom_point(aes(x = SectorID, y = `percPlantable`, 
                   color = "Plantable", shape = "Plantable" ), size = pt_size) +
    # Plot Formatting
    theme_bw() + coord_flip() +
    scale_x_discrete(limits=rev) + 
    scale_color_manual(values = c(tree_col, shrub_col, plantable_col),
                       breaks = c("Tree", "Shrub", "Plantable"),
                       guide = guide_legend(),
                       name = "Coverage") +
    scale_shape_manual(values = c(17, 16, 15),
                       breaks = c("Tree", "Shrub", "Plantable"),
                       guide = guide_legend(),
                       name = "Coverage") +
    theme(legend.position = "bottom") +
    labs(x = "", y = "Percent",
         title = "Mean plot percent coverage for each sector")
)  
  
```

Trees are generaly the highest percent coverage on average among plots for each sector. The San Carlos sector had slightly less percent coverage by trees than shrubs.  


# Trees data from i-Tree  


i-Tree tree species names were not entirely consistent (a mix of either common or scientific names first followed by the other in parentheses). We used a key table that holds the tree names organized in four columns (i-Tree name entry, Genus, Species, Common) that can be joined to the original i-Tree names and allows us to maintain the original data alongside an organized set of scientific and common name components.  

```{r message = FALSE}
key_tree_names <- read_csv(file = "../data/itree/key_tree_names.csv") |>
  select(Common, Genus, Species, iTree) |> arrange(Common)

DT::datatable(key_tree_names, rownames = FALSE, 
              colnames = c("i-Tree name entry" = "iTree"),
              caption = "Table 2. Naming key for tree species list.")
```



## Tree species by sector  

All tree species are listed alphabetically by common name with the count for each species within each sector.  

```{r}
tree_species_df <- select(itree_tree_df, SectorID, Common, Genus, Species) |> 
  reframe(.by = c("SectorID", "Common"), count = n()) |>
  pivot_wider(names_from = SectorID, values_from = count) |>
  left_join(key_tree_names, by = "Common") |>
  select(Common, Genus, Species, CN, ZC, GA, SC) |>
  arrange(Common, Genus, Species)

# print navigable table in html output
DT::datatable(tree_species_df, rownames = FALSE, 
              caption = "Table 3. Tree species list and number found in each sector.")
```



# Shrubs data from i-Tree  

i-Tree shrub species names were not entirely consistent (a mix of either common or scientific names first followed by the other in parentheses). We used a key table that holds the shrub names organized in four columns (i-Tree name entry, Genus, Species, Common) that can be joined to the original i-Tree names and allows us to maintain the original data alongside an organized set of scientific and common name components.  

```{r message=FALSE}
key_shrub_names <- read_csv(file = "../data/itree/key_shrub_names.csv") |>
  select(Common, Genus, Species, iTree) |> arrange(Common)

DT::datatable(key_shrub_names, rownames = FALSE, 
              colnames = c("i-Tree shrub name entry" = "iTree"),
              caption = "Table 4. Naming key for shrub species list.")
```

```{r}
shrub_species_df <- select(itree_shrub_df, SectorID, Common, Genus, Species) |> 
  reframe(.by = c("SectorID", "Common"), count = n()) |>
  pivot_wider(names_from = SectorID, values_from = count) |>
  left_join(key_shrub_names, by = "Common") |>
  select(Common, Genus, Species, CN, ZC, GA, SC) |>
  arrange(Common, Genus, Species)
```


## Shrub species by sector  

All shrub species are listed alphabetically common name with the count for each species within each sector.  

```{r}
DT::datatable(shrub_species_df, rownames = FALSE, 
              caption = "Table 5. Shrub species list and number found in each sector.")
```

