---
title: "MODIS Data Exploration"
author: "Matt Fuller"
date: "2024-01-12"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: true
---

```{r setup, include=FALSE}
library(sf)
library(terra)
library(tidyverse)
library(readxl)
library(kableExtra)
library(flextable)
library(leaflet)
library(htmltools)
library(viridis)
library(DT)
library(MODISTools)
library(zoo)
library(forecast)

reflat <- 18.47169 
reflong <- -69.89199

knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '')
knitr::opts_current$get('Table')
```


<br>  

# Sector geospatial data  

## Sector polygon data  

Read in sector polygons for spatial reference of MODIS data.  

```{r message = FALSE}
sector_sf <- st_read(dsn = "../data/geospatial/SD_sector_polygons.shp") |>
  select(SectorID, SectorName) |>  # select only naming fields from input
  st_zm() |>  # remove 'Z' dimension from input since it is unnecessary
  st_transform(4326)  # transform to CRS leaflet maps use by default
```


## Sector Map  

Set up basemap tiles for various backgrounds when viewing sector polygons.  

```{r warning = FALSE}
# Basemap set up  
basemap <- leaflet() |>
  addProviderTiles("OpenStreetMap.Mapnik", group = "OpenStreetMap.Mapnik") |>
  addProviderTiles("Esri.WorldTopoMap", group = "Esri.WorldTopoMap") |>
  # addProviderTiles("Esri.WorldPhysical", group = "Esri.WorldPhysical") |>
  addProviderTiles("Esri.WorldImagery", group = "Esri.WorldImagery") |>  
  # add a layers control
  addLayersControl(baseGroups = c("OpenStreetMap.Mapnik", 
                                  "Esri.WorldTopoMap",
                                  # "Esri.WorldPhysical",
                                  "Esri.WorldImagery"),
  # position it on the topleft
  position = "topleft")
```


Add polygons to basemap and plot for visualization.  

```{r}
# create palette for sector markers
sector_cols <- colorFactor(palette = viridis(4, option = "A", end = 0.85), 
                           levels = factor(sector_sf$SectorID,
                                           levels = c("CN","GA","SC","ZC")) )
(SD_plot_leaflet <- basemap |>
    addPolygons(data = sector_sf, 
                popup = ~SectorName,
                opacity = 0.8,
                color = ~sector_cols(SectorID)) 
  )
```


<br>

# MODIS product exploration  

## Product availability  

View and search products available from MODIS satellite sensors.  

```{r}
MODIS_prods <- mt_products()

DT::datatable(MODIS_prods, rownames = FALSE, 
              colnames = c("MODIS Product Code" = "product", "Description" = "description",
                           "Image Frequency" = "frequency", 
                           "Resolution (m)" = "resolution_meters"),
              caption = "Table 1. List of MODIS satellite products and their details.")

```
<br>

Looks like the MODIS products of interest are:  
Air Temperature  
1. MOD11A2: MODIS/Terra Land Surface Temperature and Emissivity (LST) 8-Day L3 Global 1 km SIN Grid  
2. MOD21A2: MODIS/Terra Land Surface Temperature/3-Band Emissivity (LSTE) 8-Day L3 Global 1 km SIN Grid

MOD11A2 filters out extreme high temperature values that MOD21A2 does not (see: https://lpdaac.usgs.gov/products/mod21a2v061/). We will use the MOD11A2 data in further temperature analyses.

Vegetation  
1. MOD13Q1: MODIS/Terra Vegetation Indices (NDVI/EVI) 16-Day L3 Global 250m SIN Grid  
2. MCD15A3H: MODIS/Terra+Aqua Leaf Area Index/FPAR (LAI/FPAR) 4-Day L4 Global 500 m SIN Grid  

## MOD11A2: Land Surface Temperature  

```{r}
bnd_MOD11A2 <- mt_bands(product = "MOD11A2")
```

```{r}
dates_MOD11A2 <- mt_dates(product = "MOD11A2", lat = reflat, lon = reflong)
```

sd_LST

```{r cache = TRUE, message=FALSE}
sd_LST_day <- mt_subset(product = "MOD11A2",
                    lat = reflat,
                    lon =  reflong,
                    band = "LST_Day_1km",
                    start = "2000-01-01",
                    end = "2024-01-01",
                    km_lr = 15,
                    km_ab = 5,
                    site_name = "SantoDomingo",
                    progress = TRUE )

sd_LST_nght <- mt_subset(product = "MOD11A2",
                    lat = reflat,
                    lon =  reflong,
                    band = "LST_Night_1km",
                    start = "2000-01-01",
                    end = "2024-01-01",
                    km_lr = 15,
                    km_ab = 5,
                    site_name = "SantoDomingo",
                    progress = TRUE )

```


```{r}
# save raw download data
saveRDS(sd_LST_day, file = "../data/RDS/modis_airtemp_day_raw.RDS")
saveRDS(sd_LST_nght, file = "../data/RDS/modis_airtemp_night_raw.RDS")
```




join the night and day data.

```{r}
sd_LST <- sd_LST_day |>
  bind_rows(sd_LST_nght) |>
  mutate(kelvin = value * 0.02) |>
  mutate(celcius = if_else(value == 0,-999, kelvin-273.15) ) |>
  filter(celcius != -999)
```


```{r}
sd_LST_date_vec <- tibble(cal_date = unique(date(sd_LST_day$calendar_date)) ) |>
  arrange(cal_date)
sd_LST_summary <- sd_LST |>
  reframe(.by = c("band", "calendar_date"), 
          temp_c = mean(celcius)) |>
  mutate(cal_date = date(calendar_date)) |>
  #filter out days with only day or night temps due to clouds or other issues
  group_by(cal_date) |>
  mutate(n_count = n()) |> ungroup() |>
  filter(n_count == 2) |>
  # calculate the "mean daily temp" using average btwn day and night temps
  pivot_wider(id_cols = cal_date, names_from = band, values_from = temp_c) |>
  rowwise() |>
  mutate(daily_TC = mean(LST_Day_1km, LST_Night_1km)) |> ungroup() |>
  select(cal_date, daily_TC) |>
  right_join(sd_LST_date_vec, by = "cal_date")
```


```{r}
ggplot(data = sd_LST_summary,
       aes(x = cal_date, y = daily_TC)) +
  # geom_point() + 
  geom_line() +
  geom_smooth() +
  theme_bw() +
  labs(x = "Date", y = "Temperature (°C)")
```

Condense temperature time series from every 8-days to mean monthly values for time series analysis and decomposition.  

```{r}
temp_zoo <- zoo(x = sd_LST_summary$daily_TC, 
               order.by = sd_LST_summary$cal_date)
plot.zoo(temp_zoo)

```

```{r}
temp_zoo_naapprox <- na.approx(temp_zoo) |> fortify.zoo() |>
  rename(cal_date = "Index", temp_C = "na.approx(temp_zoo)")
  

temp_month_df <- temp_zoo_naapprox |>
  mutate(cal_year = year(cal_date),
         cal_month = month(cal_date)) |>
  reframe(.by = c("cal_year", "cal_month"),
          avg_month_TC = mean(temp_C))

temp_month_ts <- ts(temp_month_df$avg_month_TC, frequency = 12,
                    start = c(2000,2), end = c(2024, 1))
```




```{r}
# trend analysis
temp_month_trend <- ma(temp_month_ts, order = 12, centre = TRUE)
plot.ts(temp_month_ts)
lines(temp_month_trend, col = "red")
```

```{r}
# detrend the time series for seasonality 
temp_month_detrend <- temp_month_ts - temp_month_trend
plot(as.ts(temp_month_detrend))
```


```{r}
# mean seasonality
temp_month_mtxseas <- t(matrix(data = temp_month_detrend, nrow = 12))
temp_month_avgseas <- colMeans(temp_month_mtxseas, na.rm = TRUE)
plot(as.ts(rep(temp_month_avgseas,24)))
```

```{r}
# examine random noise left over
temp_month_random <- temp_month_ts - temp_month_trend - temp_month_avgseas
plot(as.ts(temp_month_random))
```

```{r}
# nearly identical when reconstructing from the various components of the ts
plot.ts(temp_month_ts)
plot(as.ts((temp_month_trend + temp_month_avgseas + temp_month_random)))
```


```{r}
# use decompose function for trend analysis
temp_month_decomp <- decompose(temp_month_ts, type = "additive")

plot(as.ts(temp_month_decomp$seasonal))
plot(as.ts(temp_month_decomp$trend))
plot(as.ts(temp_month_decomp$random))
plot(temp_month_decomp)
```

```{r}
# save temperature data for city as a whole
saveRDS(temp_zoo_naapprox, file = "../data/RDS/modis_airtempC_8day.RDS")
saveRDS(temp_month_ts, file = "../data/RDS/modis_airtempC_monthly_ts.RDS")
```



## MOD13Q1: NDVI/EVI    

MOD13Q1: MODIS/Terra Vegetation Indices (NDVI/EVI) 16-Day L3 Global 250m SIN Grid  

```{r}
bnd_MOD13Q1 <- mt_bands(product = "MOD13Q1")
```

```{r}
dates_MOD13Q1 <- mt_dates(product = "MOD13Q1", lat = reflat, lon = reflong)
```



```{r cache = TRUE, message=FALSE}
sd_NDVI_raw <- mt_subset(product = "MOD13Q1",
                    lat = reflat,
                    lon =  reflong,
                    band = "250m_16_days_NDVI",
                    start = "2000-01-01",
                    end = "2024-01-01",
                    km_lr = 5,
                    km_ab = 5,
                    site_name = "SantoDomingo",
                    progress = TRUE )

sd_EVI_raw <- mt_subset(product = "MOD13Q1",
                    lat = reflat,
                    lon =  reflong,
                    band = "250m_16_days_EVI",
                    start = "2000-01-01",
                    end = "2024-01-01",
                    km_lr = 5,
                    km_ab = 5,
                    site_name = "SantoDomingo",
                    progress = TRUE )

```


```{r}
# save raw download data
saveRDS(sd_NDVI_raw, file = "../data/RDS/modis_NDVI_raw.RDS")
saveRDS(sd_EVI_raw, file = "../data/RDS/modis_EVI_raw.RDS")
```



```{r}
# use scale factors to revert to actual NDVI/EVI values
sd_NDVI <- sd_NDVI_raw |>
  mutate(NDVI = value * 0.0001) |>
  mutate(cal_date = date(calendar_date))
sd_EVI <- sd_EVI_raw |>
  mutate(EVI = value * 0.0001)  |>
  mutate(cal_date = date(calendar_date))
```


```{r}
sd_NDVI_summary <- sd_NDVI |>
  reframe(.by = c("calendar_date"), 
          mean_val = mean(NDVI)) |>
  mutate(cal_date = date(calendar_date), 
         parameter = "NDVI")  

sd_EVI_summary <- sd_EVI |>
  reframe(.by = c("calendar_date"), 
          mean_val = mean(EVI)) |>
  mutate(cal_date = date(calendar_date), 
         parameter = "EVI") 

sd_NDVIEVI_sum_df <- bind_rows(sd_NDVI_summary,sd_EVI_summary)
```



```{r}
NDVI_zoo <- zoo(x = sd_NDVI_summary$mean_val, 
               order.by = sd_NDVI_summary$cal_date)

NDVI_zoo_naapprox <- na.approx(NDVI_zoo) |> fortify.zoo() |>
  rename(cal_date = "Index", mean_val = "na.approx(NDVI_zoo)")
  

NDVI_month_df <- NDVI_zoo_naapprox |>
  mutate(cal_year = year(cal_date),
         cal_month = month(cal_date)) |>
  reframe(.by = c("cal_year", "cal_month"),
          avg_month_NDVI = mean(mean_val)) |>
  mutate(cal_date = date(paste0(cal_year,"-",
                                str_pad(cal_month, width=2, side="left", pad="0"),
                                "-15")))

NDVI_month_ts <- ts(NDVI_month_df$avg_month_NDVI, frequency = 12,
                    start = c(2000,2), end = c(2024, 1))
```




```{r}
EVI_zoo <- zoo(x = sd_EVI_summary$mean_val, 
               order.by = sd_EVI_summary$cal_date)

EVI_zoo_naapprox <- na.approx(EVI_zoo) |> fortify.zoo() |>
  rename(cal_date = "Index", mean_val = "na.approx(EVI_zoo)")
  

EVI_month_df <- EVI_zoo_naapprox |>
  mutate(cal_year = year(cal_date),
         cal_month = month(cal_date)) |>
  reframe(.by = c("cal_year", "cal_month"),
          avg_month_EVI = mean(mean_val)) |>
  mutate(cal_date = date(paste0(cal_year,"-",
                                str_pad(cal_month, width=2, side="left", pad="0"),
                                "-15")))

EVI_month_ts <- ts(EVI_month_df$avg_month_EVI, frequency = 12,
                    start = c(2000,2), end = c(2024, 1))
```




```{r}
ggplot(data = sd_NDVI_summary,
       aes(x = cal_date, y = mean_val)) +
  geom_line() +
  geom_line(data = NDVI_month_df,
            aes(x = cal_date, y = avg_month_NDVI), color = "red") +
  geom_smooth() +
  theme_bw() +
  labs(x = "Date", y = "NDVI", title = "Normalized Difference Vegetation Index (NDVI)")
```



```{r}
ggplot(data = sd_EVI_summary,
       aes(x = cal_date, y = mean_val)) +
  geom_line(data = EVI_month_df,
            aes(x = cal_date, y = avg_month_EVI), color = "red") +
  geom_line() +
  geom_smooth() +
  theme_bw() +
  labs(x = "Date", y = "EVI", title = "Enhanced Vegetation Index (EVI)")
```


```{r}
# save NDVI/EVI data for city as a whole
saveRDS(NDVI_zoo_naapprox, file = "../data/RDS/modis_NDVI_16day.RDS")
saveRDS(NDVI_month_ts, file = "../data/RDS/modis_NDVI_monthly_ts.RDS")

saveRDS(EVI_zoo_naapprox, file = "../data/RDS/modis_EVI_16day.RDS")
saveRDS(EVI_month_ts, file = "../data/RDS/modis_EVI_monthly_ts.RDS")
```




## MCD15A3H: LAI/FPAR    

MCD15A3H: MODIS/Terra+Aqua Leaf Area Index/FPAR (LAI/FPAR) 4-Day L4 Global 500 m SIN Grid

```{r}
bnd_MCD15A3H <- mt_bands(product = "MCD15A3H")
```

```{r}
dates_MCD15A3H <- mt_dates(product = "MCD15A3H", lat = reflat, lon = reflong)
```



```{r cache = TRUE, message=FALSE}
sd_LAI_raw <- mt_subset(product = "MCD15A3H",
                    lat = reflat,
                    lon =  reflong,
                    band = "Lai_500m",
                    start = "2000-01-01",
                    end = "2024-01-01",
                    km_lr = 5,
                    km_ab = 5,
                    site_name = "SantoDomingo",
                    progress = TRUE )

sd_FPAR_raw <- mt_subset(product = "MCD15A3H",
                    lat = reflat,
                    lon =  reflong,
                    band = "Fpar_500m",
                    start = "2000-01-01",
                    end = "2024-01-01",
                    km_lr = 5,
                    km_ab = 5,
                    site_name = "SantoDomingo",
                    progress = TRUE )

```

```{r}
# save raw download data
saveRDS(sd_LAI_raw, file = "../data/RDS/modis_LAI_raw.RDS")
saveRDS(sd_FPAR_raw, file = "../data/RDS/modis_FPAR_raw.RDS")
```



```{r}
# use scale factors to revert to actual NDVI/EVI values
sd_LAI <- sd_LAI_raw |>
  mutate(LAI = value * 0.1) |>
  mutate(cal_date = date(calendar_date))
sd_FPAR <- sd_FPAR_raw |>
  mutate(FPAR = value * 0.01)  |>
  mutate(cal_date = date(calendar_date))
```


```{r}
sd_LAI_summary <- sd_LAI |>
  reframe(.by = c("calendar_date"), 
          mean_val = mean(LAI)) |>
  mutate(cal_date = date(calendar_date), 
         parameter = "LAI")  

sd_FPAR_summary <- sd_FPAR |>
  reframe(.by = c("calendar_date"), 
          mean_val = mean(FPAR)) |>
  mutate(cal_date = date(calendar_date), 
         parameter = "FPAR") 

sd_LAIFPAR_sum_df <- bind_rows(sd_LAI_summary,sd_FPAR_summary)
```

```{r}
LAI_zoo <- zoo(x = sd_LAI_summary$mean_val, 
               order.by = sd_LAI_summary$cal_date)

LAI_zoo_naapprox <- na.approx(LAI_zoo) |> fortify.zoo() |>
  rename(cal_date = "Index", mean_val = "na.approx(LAI_zoo)")
  

LAI_month_df <- LAI_zoo_naapprox |>
  mutate(cal_year = year(cal_date),
         cal_month = month(cal_date)) |>
  reframe(.by = c("cal_year", "cal_month"),
          avg_month_LAI = mean(mean_val)) |>
  mutate(cal_date = date(paste0(cal_year,"-",
                                str_pad(cal_month, width=2, side="left", pad="0"),
                                "-15")))

LAI_month_ts <- ts(LAI_month_df$avg_month_LAI, frequency = 12,
                    start = c(2000,2), end = c(2024, 1))
```


```{r}
FPAR_zoo <- zoo(x = sd_FPAR_summary$mean_val, 
               order.by = sd_FPAR_summary$cal_date)

FPAR_zoo_naapprox <- na.approx(FPAR_zoo) |> fortify.zoo() |>
  rename(cal_date = "Index", mean_val = "na.approx(FPAR_zoo)")
  

FPAR_month_df <- FPAR_zoo_naapprox |>
  mutate(cal_year = year(cal_date),
         cal_month = month(cal_date)) |>
  reframe(.by = c("cal_year", "cal_month"),
          avg_month_FPAR = mean(mean_val)) |>
  mutate(cal_date = date(paste0(cal_year,"-",
                                str_pad(cal_month, width=2, side="left", pad="0"),
                                "-15")))

FPAR_month_ts <- ts(FPAR_month_df$avg_month_FPAR, frequency = 12,
                    start = c(2000,2), end = c(2024, 1))
```


```{r}
ggplot(data = sd_LAI_summary,
       aes(x = cal_date, y = mean_val)) +
  geom_line() +
  geom_smooth() +
  geom_line(data = LAI_month_df,
            aes(x = cal_date, y = avg_month_LAI), color = "red") +
  geom_smooth(data = LAI_month_df,
            aes(x = cal_date, y = avg_month_LAI), color = 'yellow') +
  theme_bw() +
  labs(x = "Date", y = "LAI (m^2^/m^2^)", title = "Leaf area index (LAI)")
```



```{r}
ggplot(data = sd_FPAR_summary,
       aes(x = cal_date, y = mean_val)) +
  geom_line() +
  geom_line(data = FPAR_month_df,
            aes(x = cal_date, y = avg_month_FPAR), color = "red") +
  geom_smooth() +
  theme_bw() +
  labs(x = "Date", y = "FPAR", title = "Fraction of photosynthetically active radiation (FPAR)")
```



```{r}
# save LAI/FPAR data for city as a whole
saveRDS(LAI_zoo_naapprox, file = "../data/RDS/modis_LAI_4day.RDS")
saveRDS(LAI_month_ts, file = "../data/RDS/modis_LAI_monthly_ts.RDS")

saveRDS(FPAR_zoo_naapprox, file = "../data/RDS/modis_FPAR_4day.RDS")
saveRDS(FPAR_month_ts, file = "../data/RDS/modis_FPAR_monthly_ts.RDS")
```


