---
title: "Developing urban-subruban gradient"
author: "Matt Fuller"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: yes
    number_sections: yes
    toc_float: yes
    code_folding: hide
  pdf_document:
    toc: yes
---

Initial set up and R package loading.  

```{r setup, message=FALSE, warning = FALSE}
library(tidyverse)
library(readxl)
library(data.table)
library(dtplyr)
library(foreign)
library(terra)
library(tidyterra)
library(leaflet)
library(basemaps)
library(ggspatial)
library(sf)
# devtools::install_github("yutannihilation/ggsflabel")
library(ggsflabel)
library(viridisLite)
library(viridis)
library(factoextra)

knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '')
```


<br>  

# Read in data sets  


```{r read_in_bird_data, message=FALSE}
## Bird data ##

birditree_sitejoin_key <- readRDS(file = "../data/iTree_Bird_site_key.RDS")

# Bird site data
bird_dist_m_to_shore <- 
  read_csv(file = "../data/bird_SiteID_with_dist_m_to_watercoast.csv") |> 
  select(SiteID, h2o_dist_m)

bird_dist_m_to_shore$waterbody_dist <- 
  scales::rescale(bird_dist_m_to_shore$h2o_dist_m,
                  to = c(0,1),
                  from = range(bird_dist_m_to_shore$h2o_dist_m, 
                               na.rm = TRUE, finite = TRUE))
  
bird_site_data_raw <- readRDS(file = "../data/RDS/bird_sitevisit_data.RDS")
bird_site_data <- bird_site_data_raw |>
  filter(!(SiteID %in% c(paste0("ZCA0",1:7),"SC15", "GA999"))) |>
  left_join(bird_dist_m_to_shore, by = "SiteID")

# i-Tree plot data
itree_data <- readRDS(file = "../data/itree/itree_plot_data_organized.RDS") |>
  right_join(select(birditree_sitejoin_key, SectorName, SiteID, itreeplotID), 
             by = c("itreeplotID")) |>
  filter(!(SiteID %in% c(paste0("ZCA0",1:7),"SC15", "GA999"))) |>
  select(SectorName, SiteID, percTree, percShrub, percPlantable) |>
  mutate(across(where(is.numeric), \(x) (x/100) ) )
```



```{r read_ing_geospatial_data, message = FALSE}
## Geospatial data ##

# spatial data for sites
bird_siteID_sf <- birditree_sitejoin_key |>
  filter(!is.na(PlotNum)) |>
  sf::st_as_sf(coords = c("long", "lat"), crs = st_crs(4326))

# create 50m-radius buffer polygons for land use analysis at each site
site_050m_buffer_polygons <- st_buffer(x = bird_siteID_sf, dist = 50)

# read in raster with gray, green, and blue infrastructure and reclassify by
# combining non-residential and residential clases of developed, trees, & grass
lc_raster <- 
  terra::rast(x = "../data/geospatial/lc_resi_25march2019_waterupdate.tif") #|>
  # terra::classify(rcl = matrix(data = c(0,1,2,4,5,10,11,12,14,15,
  #                                       0,1,2,4,5, 0, 1, 2, 4, 5), ncol = 2))

# set the category names for each class of infrastructure
# levels(lc_raster) <- 
#   data.frame(ID = c(0,1,2,4,5), 
#              category = c("Developed","Tree","Grass","Water","No Data"))

# grab crs for lulc raster
lulc_crs <- terra::crs(lc_raster)


# 
# site_050m_buffer_polygons <-
#   st_read(dsn = "../data/geospatial/Site_050m_buffer_polygons.shp",
#           quiet = TRUE) |>
#   filter(!(SiteID %in% c(paste0("ZCA0",1:7),"SC15", "GA999")))

sector_polygons <- 
  st_read(dsn = "../data/geospatial/SD_sector_polygons.shp", quiet = TRUE)  
```


Land use data published originally at: https://doi.org/10.1016/j.ufug.2021.127125 for USAID in partnership with USDA Forest Service.  


# Mapping of sectors, sites, and buffers  

## Interactive map of sectors and sites

Zoomable map for exploring site/sector areas using background maps with streets or aerial imagery.  

```{r leaflet_basemap, warning = FALSE}
# Set up basemap tiles for various background maps
basemap <- leaflet() |>
  # add background maps
  addProviderTiles("Esri.WorldImagery", group = "Esri.WorldImagery") |>  
  addProviderTiles("Esri.WorldTopoMap", group = "Esri.WorldTopoMap") |>
  # add a layers control
  addLayersControl(baseGroups = c("Esri.WorldImagery",
                                  "Esri.WorldTopoMap"),
  # position background map selection icon in the top-left part of map
  position = "topleft")
```


```{r leaflet_interactive_map, warning = FALSE, fig.cap="**Figure 1.** Sites representing those that are matched between bird and i-Tree data across four sectors (Ciudad Nueva: black; Gazcue: purple; San Carlos: pink; Zona Colonial: orange)."}
# create palette for sector markers
sector_cols <- colorFactor(palette = viridis(4, option = "A", end = 0.85), 
                           levels = factor(bird_siteID_sf$SectorName,
                                           levels = c("Ciudad Nueva","Gazcue",
                                                      "San Carlos","Zona Colonial")) )

# Map the plots for each sector.  
(SD_plot_leaflet <- basemap |>
    addPolygons(data = site_050m_buffer_polygons) |>
    addCircleMarkers(data = bird_siteID_sf,
                     popup = ~paste0("Sector: ", SectorName, "<br>","Site ID: ", SiteID), 
                     radius = 4, opacity = 1,
                     color = ~ sector_cols(SectorName) )
)
```

<br>  


## Static maps of sectors and sites  


```{r ggplot_basemap_development, message = FALSE, warning=FALSE}
# add basemap for ggplot map
map_extent <- st_bbox(sector_polygons, crs = 3857)
map_bbox_buffer <- 100
map_extent[1] <- map_extent[1] - map_bbox_buffer
map_extent[3] <- map_extent[3] + map_bbox_buffer
map_extent[2] <- map_extent[2] - map_bbox_buffer
map_extent[4] <- map_extent[4] + map_bbox_buffer

basemap_ggplot <- ggplot()

# base ggplot for scenario maps
basemap_ggplot <- 
  basemap_ggplot(ext = map_extent, verbose = FALSE,
                 map_service = "carto", 
                 map_type = "light_no_labels") +
  
  geom_sf(data = sector_polygons, color = "red", fill = NA, linewidth = 0.8) +
  geom_sf_label(data = sector_polygons, aes(label = SectorName), size = 2.5,
                nudge_x = c(-550,150,225,450),  # SC, GA, CN, ZC
                nudge_y = c(550,600,-500,-100)) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        legend.background = element_blank()) +
  coord_sf(crs = 3857, 
           xlim = c(map_extent[1], map_extent[3]),
           ylim = c(map_extent[2], map_extent[4]),
           expand = FALSE) +
  # ggspatial additions of north arrow and scalebar
  annotation_north_arrow(location = "br", which_north = "true",
                         pad_x = unit(0.1, "in"),pad_y = unit(0.3, "in"),
                         style = north_arrow_nautical(fill = c("grey40", "white"),
                                                      line_col = "grey20",
                                                      text_family = "ArcherPro Book")) +
  annotation_scale(location = "br",bar_cols = c("grey60", "white"),
                   text_family = "ArcherPro Book")  +
  labs(x = "", y = "")

# basemap_ggplot
```


### Sectors and sites map  

```{r site_map_ggplot, warning = FALSE, message = FALSE, fig.width = 6.5, fig.height = 4.5}
(site_map_ggplot <- basemap_ggplot +
   geom_sf(data = bird_siteID_sf) +
   
   ggsflabel::geom_sf_text_repel(data = bird_siteID_sf, aes(label = SiteID),
                                  size = 2) +
   coord_sf(crs = 3857, 
           xlim = c(map_extent[1], map_extent[3]),
           ylim = c(map_extent[2], map_extent[4]),
           expand = FALSE) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        legend.background = element_blank()) +
   labs(title = "Sites within four sectors of study system in Santo Domingo" )
)

ggsave(plot = site_map_ggplot, width = 6.5, height = 4.5, dpi = 600, units = "in",
       filename = "../figs/Site_map_with_four_sectors.png")
```


### Site 50 m buffers mapped  

```{r 050m_buffer_areas_for_sites_map, warning = FALSE, message = FALSE, fig.width = 6.5, fig.height = 4.5}
(site_050mbuffer_map_ggplot <- basemap_ggplot +
   geom_sf(data = site_050m_buffer_polygons, alpha = 0.2, color = "blue", fill = "skyblue") +
   
   # geom_sf(data = site_050m_st_buffers, alpha = 0.2, color = "blue", fill = "skyblue") +
   
   geom_sf(data = bird_siteID_sf) +
   
   ggsflabel::geom_sf_text_repel(data = bird_siteID_sf, aes(label = SiteID),
                                  size = 2) +
   coord_sf(crs = 3857, 
           xlim = c(map_extent[1], map_extent[3]),
           ylim = c(map_extent[2], map_extent[4]),
           expand = FALSE) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        legend.background = element_blank()) +
   labs(title = "50 m site buffers for each site for zonal statistics" )
)

ggsave(plot = site_050mbuffer_map_ggplot, width = 6.5, height = 4.5, dpi = 600, units = "in",
       filename = "../figs/Site_map_with_four_sectors_w50m_buffers.png")
```


# Zonal statistics for land use calculations  

<br>  

## Sector-based zonal statistics {.tabset}  

Iterate through sector polygons to generate land use zonal statistics for each.  

### Sector land use analysis code  

See section tabs for tables, bar chart, and maps.  

```{r sector_lulc_zona_stats_calculations, fig.height=4, fig.width=6.5}
sector_polygons_lulc <- sector_polygons  |>
  st_transform(crs = lulc_crs)

sector_zonal_stats_df <- data.frame()

for (idx in 1:nrow(sector_polygons_lulc)){
  polygon_geom <- sector_polygons_lulc$geometry[idx] |> vect() 
  
  polygon_name <- sector_polygons_lulc$SectorName[idx]
  
  polygon_raster <- terra::crop(x = lc_raster, y = polygon_geom, mask = TRUE)
  
  # plot(polygon_raster, main = polygon_name, type = "classes",
  #      col = c("gray30","burlywood4","green4","navyblue","red4",
  #              "gray80","burlywood1","green1", "dodgerblue3","red1"),
  #      all_levels = TRUE)
  
  cell_area_m2 <- res(polygon_raster)[1]*res(polygon_raster)[2]
  
  zonal_stats <- terra::freq(x = polygon_raster) |>
    mutate(ID = polygon_name,
           area_m2 = count * cell_area_m2) |>
    mutate(area_ha = area_m2*0.0001) |>
    select(ID, value, area_m2, area_ha, count)
  
  sector_zonal_stats_df <- bind_rows(sector_zonal_stats_df,
                                     zonal_stats)
}
```



### Table (with residential split)

```{r sector_wressplit_table}
sector_zonal_stats_df |>
  mutate(.by = ID, total_ha = sum(area_ha, na.rm = TRUE)) |> ungroup() |>
  mutate(area_prop = area_ha/total_ha) |>
  pivot_wider(id_cols = ID, names_from = value, values_from = area_prop) %>%
  mutate(across(where(is.numeric), \(x) round(x, 2) ) ) |>
  DT::datatable(rownames = FALSE, 
                caption = "Table 2. Land use summary by site for 50 m buffer areas.")
```

### Table (without residential split)

```{r sector_woressplit_table}
sector_zonal_stats_df |>
  mutate(.by = ID, total_ha = sum(area_ha, na.rm = TRUE)) |> ungroup() |>
  rowwise() |>
  mutate(area_prop = area_ha/total_ha) |> ungroup() |>
  pivot_wider(id_cols = ID, names_from = value, values_from = area_prop) %>%
  rowwise() |>
  mutate(Develop = sum(`Develop, non-residential`,`Develop, residential`,na.rm = TRUE),
         Tree = sum(`Tree, non-residential`,`Tree, residential`,na.rm = TRUE),
         Grass = sum(`Grass, non-residential`,`Grass, residential`,na.rm = TRUE),
         Water = `Water, non-residential`) |> select(-contains(",")) |>
  mutate(across(where(is.numeric), \(x) round(x, 2) ) ) |>
  DT::datatable(rownames = FALSE, 
                caption = "Table 2. Land use summary by site for 50 m buffer areas.")
```




### Bar chart for land use proportions  


```{r sector_lulc_bar_chart, warning = FALSE, message = FALSE, fig.height = 2, fig.width = 6.5}
sector_zonal_stats_df |>
   mutate(.by = ID, total_ha = sum(area_ha, na.rm = TRUE)) |> ungroup() |>
  rowwise() |>
  mutate(area_prop = area_ha/total_ha) |> ungroup() |>
  pivot_wider(id_cols = ID, names_from = value, values_from = area_prop) %>%
  rowwise() |>
  mutate(Develop = sum(`Develop, non-residential`,`Develop, residential`,na.rm = TRUE),
         Tree = sum(`Tree, non-residential`,`Tree, residential`,na.rm = TRUE),
         Grass = sum(`Grass, non-residential`,`Grass, residential`,na.rm = TRUE),
         Water = `Water, non-residential`) |> select(-contains(",")) |>
  mutate(across(where(is.numeric), \(x) round(x, 2) ) ) |> ungroup() |>
  mutate(ID = factor(ID, levels = unique(ID[order(Develop, Tree, Grass, Water)]), 
                     ordered = TRUE) ) |>
  pivot_longer(cols = !ID, names_to = "landuse", values_to = "Proportion") |>
  mutate(landuse = factor(landuse, 
                          levels = c("Develop","Tree","Grass","Water"))) |>
  
  ggplot(aes(x = ID, y = Proportion)) +
  geom_col(aes(fill = landuse)) +
  theme_bw() +
  scale_fill_manual(breaks = c("Develop","Tree","Grass","Water"),
                    values = c("gray30","burlywood4","green4","navyblue")) +
  labs(x = "", fill = "Land use",
       title = "Sector land use coverage") +
  theme(legend.position = "none", 
        axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45)) +
  coord_flip()+
  facet_wrap(~landuse, ncol = 4)
```

### Sector land use maps  

```{r sector_lulc_zona_stats_maps, fig.height=4, fig.width=6.5}
for (idx in 1:nrow(sector_polygons_lulc)){
  polygon_geom <- sector_polygons_lulc$geometry[idx] |> vect() 
  
  polygon_name <- sector_polygons_lulc$SectorName[idx]
  
  polygon_raster <- terra::crop(x = lc_raster, y = polygon_geom, mask = TRUE)
  
  plot(polygon_raster, main = polygon_name, type = "classes",
       col = c("gray30","burlywood4","green4","navyblue","red4",
               "gray80","burlywood1","green1", "dodgerblue3","red1"),
       all_levels = TRUE)
}
```

<br>  

## Site-based zonal statistics {.tabset}  

Iterate through polygons for site 50 m radius areas to tabulate the areas of developed (gray), trees/grass (green), and water (blue) land covers for residential and non-residential coverages.   

### 50 m buffer site landuse analysis code  

See section tabs for tables, bar chart, and maps.  

```{r site050mbuffer_lulc_zona_stats_calculations, fig.height=4, fig.width=6.5, cache = TRUE}
# transform coordinate system to match lulc raster
site_050m_buffer_polygons_lulc <- site_050m_buffer_polygons |>
  st_transform(crs = lulc_crs)

# create data frame to hold each polygon's zonal stats data
site050m_zonal_stats_df <- data.frame()

for (idx in 1:nrow(site_050m_buffer_polygons_lulc)){
  polygon_geom <- site_050m_buffer_polygons_lulc$geometry[idx] |> vect() 
  
  polygon_name <- site_050m_buffer_polygons_lulc$SiteID[idx]
  
  polygon_raster <- terra::crop(x = lc_raster, y = polygon_geom, mask = TRUE)
  
  # plot(polygon_raster, main = polygon_name, type = "classes",
  #      col = c("gray30","burlywood4","green4","navyblue","red4",
  #              "gray80","burlywood1","green1", "dodgerblue3","red1"),
  #      all_levels = TRUE)
  
  cell_area_m2 <- res(polygon_raster)[1]*res(polygon_raster)[2]
  
  zonal_stats <- terra::freq(x = polygon_raster) |>
    mutate(ID = polygon_name,
           area_m2 = count * cell_area_m2) |>
    mutate(area_ha = area_m2*0.0001) |>
    select(ID, value, area_m2, area_ha, count)
  
  site050m_zonal_stats_df <- bind_rows(site050m_zonal_stats_df,
                                     zonal_stats)
}
```

### Table (with residential split)

```{r site050mbuffer_wressplit_table}
site050m_zonal_stats_ungroupedcategories_df <- site050m_zonal_stats_df |>
  filter(!(value %in% c("Water, non-residential"))) |>
  mutate(.by = ID, total_ha = sum(area_ha, na.rm = TRUE)) |> ungroup() |>
  rowwise() |>
  mutate(area_prop = area_ha/total_ha) |> ungroup() |>
  pivot_wider(id_cols = ID, names_from = value, 
              values_from = area_prop, values_fill = 0)
  
# save data for future analysis
saveRDS(site050m_zonal_stats_ungroupedcategories_df,
        file = "../data/RDS/site050m_zonal_stats_ungroupedcategories_df.RDS")

DT::datatable(data = mutate(site050m_zonal_stats_ungroupedcategories_df,
                            across(where(is.numeric), \(x) round(x, 2) ) ),
                rownames = FALSE, 
                caption = "Table 2. Land use summary by site for 50 m buffer areas.")
```

### Table (without residential split)

```{r site050mbuffer_woressplit_table}
site050m_zonal_stats_groupedcategories_df <- site050m_zonal_stats_df |>
  filter(!(value %in% c("Water, non-residential"))) |>
  mutate(.by = ID, total_ha = sum(area_ha, na.rm = TRUE)) |> ungroup() |>
  rowwise() |>
  mutate(area_prop = area_ha/total_ha) |> ungroup() |>
  pivot_wider(id_cols = ID, names_from = value, 
              values_from = area_prop, values_fill = 0) %>%
  rowwise() |>
  mutate(Develop = sum(`Develop, non-residential`,`Develop, residential`,na.rm = TRUE),
         Tree = sum(`Tree, non-residential`,`Tree, residential`,na.rm = TRUE),
         # Water = `Water, non-residential`,
         Grass = sum(`Grass, non-residential`,`Grass, residential`,na.rm = TRUE)
         ) |> select(-contains(","))
  

# save data for future analysis
saveRDS(site050m_zonal_stats_groupedcategories_df,
        file = "../data/RDS/site050m_zonal_stats_groupedcategories_df.RDS")


DT::datatable(data = mutate(site050m_zonal_stats_groupedcategories_df,
                            across(where(is.numeric), \(x) round(x, 2) ) ),
              rownames = FALSE, 
              caption = "Table 2. Land use summary by site for 50 m buffer areas.")
```




### Bar chart for land use proportions  


```{r site050mbuffer_lulc_bar_chart, warning = FALSE, message = FALSE, fig.height = 8.5, fig.width = 6.5}
site050m_zonal_stats_df |>
  filter(!(value %in% c("Water, non-residential"))) |>
  mutate(.by = ID, total_ha = sum(area_ha, na.rm = TRUE)) |> ungroup() |>
  rowwise() |>
  mutate(area_prop = area_ha/total_ha) |> ungroup() |>
  pivot_wider(id_cols = ID, names_from = value, values_from = area_prop) %>%
  rowwise() |>
  mutate(Develop = sum(`Develop, non-residential`,`Develop, residential`,na.rm = TRUE),
         Tree = sum(`Tree, non-residential`,`Tree, residential`,na.rm = TRUE),
         # Water = `Water, non-residential`,
         Grass = sum(`Grass, non-residential`,`Grass, residential`,na.rm = TRUE)
         ) |> select(-contains(",")) |>
  mutate(across(where(is.numeric), \(x) round(x, 2) ) ) |> ungroup() |>
  # rename(SiteID = "ID") |>
  mutate(ID = factor(ID, levels = unique(ID[order(Develop, Tree, Grass#, Water
                                                  )]), 
                     ordered = TRUE) ) |>
  pivot_longer(cols = !ID, names_to = "landuse", values_to = "Proportion") |>
  mutate(landuse = factor(landuse, 
                          levels = c("Develop","Tree","Grass","Water"))) |>
  
  ggplot(aes(x = ID, y = Proportion)) +
  geom_col(aes(fill = landuse)) +
  theme_bw() +
  scale_fill_manual(breaks = c("Develop","Tree","Grass","Water"),
                    values = c("gray30","burlywood4","green4","navyblue")) +
  labs(x = "", fill = "Land use",
       title = "Site land use coverage within 50 m buffers") +
  theme(legend.position = "none", 
        axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45)) +
  coord_flip()+
  facet_wrap(~landuse, ncol = 4)

ggsave(plot = last_plot(), filename = "../figs/landcover_barplot_grouped.png",
       width = 6.5, height = 8.5, units = "in", dpi = 600)
```



### 50 m buffer site landuse maps  

```{r site050mbuffer_lulc_zona_stats_maps, fig.height=4, fig.width=6.5}
for (idx in 1:nrow(site_050m_buffer_polygons_lulc)){
  polygon_geom <- site_050m_buffer_polygons_lulc$geometry[idx] |> vect() 
  
  polygon_name <- site_050m_buffer_polygons_lulc$SiteID[idx]
  
  polygon_raster <- terra::crop(x = lc_raster, y = polygon_geom, mask = TRUE)
  
  plot(polygon_raster, main = polygon_name, type = "classes",
       col = c("gray30","burlywood4","green4","navyblue","red4",
               "gray80","burlywood1","green1", "dodgerblue3","red1"),
       all_levels = TRUE)
}
```

# Rank-based urban-suburban gradient  

Rank-based gradient provides user-specified weighting to each component used for extracting a gradient between urban and suburban areas. Here we create several weighting schemes from the four land cover categories (Developed, Trees, Grass, distance to permanent water body) that include:  

1. Equal weights for each category  
2. variable weighting assignments that weight developed areas higher than than green spaces (Trees and Grass) and finally distance to water bodies. 

```{r}
rank_grdt_df <- site050m_zonal_stats_groupedcategories_df |>
  rename(SiteID = "ID") |>
  left_join(bird_dist_m_to_shore, by = "SiteID") |> ungroup() |>
  mutate(rank_dev = min_rank(desc(Develop)),
         rank_tre = min_rank(Tree),
         rank_gra = min_rank(Grass),
         rank_h2o = min_rank(desc(h2o_dist_m)) ) |>
  rowwise() |>
  # equal weights
  mutate(rank_wght_25252525 = sum(rank_dev*0.25, rank_tre*0.25, 
                              rank_gra*0.25, rank_h2o*0.25)) |> 
  # weights 50% dev, 25% tree, 15% grass, 10% water
  mutate(rank_wght_50251510 = sum(rank_dev*0.5, rank_tre*0.25, 
                              rank_gra*0.15, rank_h2o*0.1)) |> 
  # weights 50% dev, 20% tree, 15% grass, 15% water
  mutate(rank_wght_50201515 = sum(rank_dev*0.5, rank_tre*0.20, 
                              rank_gra*0.15, rank_h2o*0.15)) |> 
  # weights 45% dev, 15% tree, 15% grass, 15% water
  mutate(rank_wght_45151515 = sum(rank_dev*0.45, rank_tre*0.15, 
                              rank_gra*0.15, rank_h2o*0.15)) |> 
  # weights 70% dev, 10% tree, 10% grass, 10% water
  mutate(rank_wght_70101010 = sum(rank_dev*0.7, rank_tre*0.1, 
                              rank_gra*0.1, rank_h2o*0.1)) |> 
  # weights 40% dev, 20% tree, 20% grass, 20% water
  mutate(rank_wght_40202020 = sum(rank_dev*0.4, rank_tre*0.2, 
                              rank_gra*0.2, rank_h2o*0.2)) |> 
  # weights 30% dev, 30% tree, 30% grass, 10% water
  mutate(rank_wght_30303010 = sum(rank_dev*0.3, rank_tre*0.3, 
                              rank_gra*0.3, rank_h2o*0.1)) |> 
  # weights 30% dev, 30% tree, 20% grass, 20% water
  mutate(rank_wght_30302020 = sum(rank_dev*0.3, rank_tre*0.3, 
                              rank_gra*0.2, rank_h2o*0.2)) |> 
  ungroup() |> 
  select(SiteID, starts_with("rank_wght"))

saveRDS(rank_grdt_df, file = "../data/RDS/rankbased_urbangrdt_df.RDS")
```



# PCA-based urban-suburban gradient  

## PCA code resources  

Basic PCA analysis:
https://www.datacamp.com/tutorial/pca-analysis-r

Visualizing PCA results:
https://rpkgs.datanovia.com/factoextra/reference/fviz_pca.html


## Data preparation for PCA  

```{r}
site050m_zonal_stats_pca_df <- site050m_zonal_stats_groupedcategories_df %>%
  mutate(across(where(is.numeric), ~replace(., is.na(.), 0))) |>
  left_join(itree_data, by = c("ID" = "SiteID")) |>
  left_join(select(bird_dist_m_to_shore, SiteID, waterbody_dist), 
            by = c("ID" = "SiteID") ) |>
  mutate(SectorName = factor(SectorName, 
                             levels = c("Gazcue","Ciudad Nueva",
                                        "Zona Colonial","San Carlos"))) |>
  column_to_rownames(var = "ID") |>
  select(SectorName, everything())
```

## Run PCA  


```{r}
urban_pca <- prcomp(site050m_zonal_stats_pca_df[,-1], scale = TRUE)
```


## Visualize PCA results  

### Scree plot  


```{r}
factoextra::fviz_eig(urban_pca)
```


### Variable loadings plot


```{r fig.height = 6.5, fig.width=6.5, warning = FALSE}
(PCA_variable_loadings_plot <- 
   factoextra::fviz_pca_var(urban_pca, col.circle = NA,
                            # gradient.cols = rocket(10, direction = -1, begin = 0.5, end = 1),
                            # gradient.cols = cividis(10, direction = -1),
                            # gradient.cols = plasma(10, begin = 0, end = 0.5), 
                            col.var = "contrib") +
   theme_bw() +
   scale_color_viridis(option = "F", direction = -1, begin = 0.5, end = 1,
                       breaks = c(5.5, 20), labels = c("Low","High"), 
                       guide = "colorbar") +
   labs(title = "Suburban-urban gradient PCA: Variable loadings",
        subtitle = "Variable importance: % contribution of each variable to dimension 1 and 2 spread",
        color = "Variable Importance") +
   theme(legend.direction = "horizontal",
         # legend.direction = "vertical", 
         # legend.position = c(0.92,0.15),   # vertical lower right gradient
         legend.position = c(0.86,0.92),   # horizontal top right gradient
         legend.background = element_blank()) +
   guides(color = guide_colorbar(title.position = "top", title.hjust = 0.5))
)

ggsave(PCA_variable_loadings_plot, 
       filename = "../figs/urban_gradient_PCA_variable_loadings_plot.png",
       width = 6.5, height = 6.5, units = "in", dpi = 600)
```


### Individual sites plot  

```{r fig.height = 6.5, fig.width=6.5, warning = FALSE}
(PCA_individual_sites_sector_ellipse_plot <- 
   factoextra::fviz_pca_ind(urban_pca, label = "none",
                            habillage = site050m_zonal_stats_pca_df$SectorName, 
                            addEllipses = TRUE, ellipse.level=0.95, palette = "Dark2",
                            alpha.ind = 0.2,
                            repel = TRUE)  +
   theme_bw() +
   scale_shape_manual(values = c(15, 19, 17, 18),
                      breaks = c("Gazcue","Ciudad Nueva","Zona Colonial","San Carlos")) +
   labs(title = "Suburban-urban  gradient PCA: Sites",
        subtitle = "",
        shape = "Sector:", color = "Sector:", fill = "Sector:") +
   theme(legend.direction = "horizontal", 
         legend.position = c(-0.01,1.0275),  legend.justification = "left",
         # legend.position = c(0,0.03),  legend.justification = "left", 
         legend.background = element_blank()) 
)

ggsave(PCA_individual_sites_sector_ellipse_plot, 
       filename = "../figs/urban_gradient_PCA_individuals_sectorellipses_plot.png",
       width = 6.5, height = 6.5, units = "in", dpi = 600)

```


### Bar plot: Variable contributions to dimensions 1 and 2  

```{r}
factoextra::fviz_contrib(urban_pca, choice = "var", axes = 1:2)
```


## Extract PCA dimension values for each site  

```{r}
urban_pca_dims <- factoextra::get_pca(urban_pca, element = "ind")$coord |>
  data.frame() |>
  rownames_to_column(var = "SiteID")

saveRDS(urban_pca_dims, file = "../data/RDS/urban_gradient_pca_dims.RDS")
```


